---
title: "Portfolio Part 1: Development and Planning"
author: "Dhruv Salian"
format: html
---

## Data Description

### Data Source

For this project I will use a Spotify dataset obtained from Kaggle, specifically the "Spotify Tracks Dataset." This dataset contains information about songs on Spotify including audio features and popularity scores.
The dataset can be accessed at: https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset

### Description of Data

Each row in the dataset represents a single song. The dataset includes several variables such as:

* **track_name** (character): the name of the song
* **artists** (character): the artist(s) who performed the song
* **popularity** (numeric): a score from 0–100 representing how popular the song is
* **explicit** (logical): whether the song contains explicit content or not
* **loudness** (numeric): how loud the song is, measured in decibels
* **acousticness** (numeric): how acoustic the song sounds, from 0 (electronic) to 1 (fully acoustic)
* **danceability** (numeric): how suitable a track is for dancing
* **energy** (numeric): intensity and activity level of the song
* **valence** (numeric): positivity or mood of the song
* **tempo** (numeric): speed of the song in beats per minute
* **track_genre** (character): the genre the song belongs to

These variables include a mix of numeric, character, and logical types.

### Research Questions

1. Do explicit songs tend to be more popular than non-explicit songs on Spotify?
2. Do louder songs tend to be more popular — is there a relationship between loudness and popularity?
3. Are acoustic songs less popular than electronic-sounding ones on Spotify?

---

## Data Visualization

### What do you want your final visualizations to look like?

### What do you want your final visualizations to look like?

For the first question I plan to create a violin plot comparing the 
distribution of popularity scores between explicit and non-explicit songs. 
For the second question I plan to create a line chart showing how average 
popularity changes across four loudness groups from very quiet to very loud. 
For the third question I plan to create a horizontal bar chart comparing 
average popularity across low, medium, and high acousticness groups.

### What do you want to highlight in your visualizations?

I want to highlight whether certain song characteristics — like being explicit, being louder, or being more acoustic — are connected to how popular a song becomes on Spotify. Each visualization will focus on one characteristic and show clearly whether songs with that trait tend to perform better or worse in terms of popularity.

### What is missing from your data or needs to change?

The acousticness variable is a continuous number from 0 to 1, so I will need to group it into Low, Medium, and High categories to make comparisons easier. The explicit variable is stored as text and may need to be converted to a proper logical or factor format for clean visualization.

---

## Data Cleaning

### Do you need to reformat variables?

Yes. I will need to make sure numeric variables like loudness, acousticness, and popularity are correctly formatted. I will also convert the explicit variable from text into a proper labeled format so it reads clearly in visualizations.

### Do you need to deal with missing data?

Yes. I will check all key variables for missing values and remove any rows where important fields are blank or unavailable.

### Do you need to filter your data?

Yes. I will remove songs with a popularity score of zero since those are typically unlisted or unreleased tracks that are not actually available to listeners and would skew the results.

### Do you need to create new variables?

Yes. I will create a grouped acousticness variable with three levels — Low, Medium, and High — to make it easier to compare acoustic versus electronic songs visually.

### Do you need to join additional data?

No. The dataset already contains all the variables needed for this analysis.

### Do you need to pivot your data?

No. The dataset is already in a usable format where each row is one song.

### Do you need to summarize variables?

Yes. I will calculate average popularity scores for each group in the acousticness and explicit comparisons so the bar chart and boxplot are easy to interpret.

### What other cleaning is needed?

I will remove duplicate songs so each track only appears once in the analysis. I will also remove any columns that are not relevant to my research questions, such as track IDs and album names.

### Are there variables you can exclude?

Yes. I plan to exclude variables like track ID, album name, key, mode, time signature, speechiness, instrumentalness, liveness, and tempo since they are not directly relevant to the three research questions I am investigating.

---

## Initial Setup

```{r}
#| label: load-libraries

# Load libraries
library(tidyverse)
library(here)
```

```{r}
#| label: load-data

# Import data
spotify <- read_csv(here("data", "dataset.csv"))
```

```{r}
#| label: preview-data

# Preview data
glimpse(spotify)
summary(spotify)
```

```{r}
#| label: clean-data

# keeping only what i need for my analysis
spotify_clean <- spotify |>
  select(
    track_name,
    artists,
    track_genre,
    popularity,
    explicit,
    loudness,
    acousticness,
    danceability,
    energy,
    valence,
    tempo
  )

# removing songs with missing values and zero popularity
# zero popularity usually means the song isnt publicly available
spotify_clean <- spotify_clean |>
  filter(
    !is.na(popularity),
    !is.na(loudness),
    !is.na(acousticness),
    !is.na(explicit),
    popularity > 0
  )

# converting explicit from True/False text to proper labels
spotify_clean <- spotify_clean |>
  mutate(
    explicit = case_when(
      explicit == TRUE  | explicit == "True"  ~ "Explicit",
      explicit == FALSE | explicit == "False" ~ "Not Explicit",
      TRUE ~ "Not Explicit"
    ),
    explicit = factor(explicit, levels = c("Not Explicit", "Explicit"))
  )

# grouping acousticness into three levels so its easier to compare
spotify_clean <- spotify_clean |>
  mutate(
    acousticness_level = case_when(
      acousticness < 0.33  ~ "Low Acousticness",
      acousticness < 0.66  ~ "Medium Acousticness",
      acousticness >= 0.66 ~ "High Acousticness"
    ),
    acousticness_level = factor(
      acousticness_level,
      levels = c("Low Acousticness", "Medium Acousticness", "High Acousticness")
    )
  )

# removing duplicate songs so each track only shows up once
spotify_clean <- spotify_clean |>
  distinct(track_name, artists, .keep_all = TRUE)

# getting average popularity per acousticness group for the bar chart
acousticness_summary <- spotify_clean |>
  group_by(acousticness_level) |>
  summarize(
    avg_popularity = mean(popularity, na.rm = TRUE),
    count = n()
  ) |>
  arrange(desc(avg_popularity))

glimpse(spotify_clean)
summary(spotify_clean)
```
