---
title: "Project: What Makes a Song Popular on Spotify?"
author: "Dhruv Salian"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Data Description

For this project, I used the Spotify Tracks Dataset from [Kaggle](https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset). This dataset contains information on over 114,000 songs available on Spotify, including audio characteristics and a popularity score assigned by the platform for each song.

Each row in the dataset represents one unique song. The key variables I used in this analysis are:

- **track_name**: The name of the song
- **artists**: Who performed the song
- **popularity**: A number from 0 to 100 showing how popular the song is on Spotify — higher means more popular
- **explicit**: Whether the song contains explicit content (True or False)
- **loudness**: How loud the song is, measured in decibels — most songs fall between -50 and 0
- **acousticness**: How acoustic the song sounds, from 0 (fully electronic) to 1 (fully acoustic)
- **track_genre**: The genre the song belongs to

The three questions I set out to answer were:

1. Do explicit songs tend to be more popular than non-explicit songs?
2. Do louder songs tend to be more popular?
3. Are acoustic songs less popular than electronic-sounding ones?

---

## Data Cleaning

```{r}
#| label: load-packages

library(tidyverse)
library(here)
```

```{r}
#| label: load-data

# Load the raw Spotify dataset using the here package
spotify_raw <- read_csv(here("data", "dataset.csv"))
```

```{r}
#| label: clean-data

# Step 1: SELECT only the columns we need for this analysis
# This removes unneeded columns like track IDs, album names, and unused audio features
spotify_clean <- spotify_raw |>
  select(
    track_name,
    artists,
    track_genre,
    popularity,
    explicit,
    loudness,
    acousticness,
    danceability,
    energy,
    valence,
    tempo
  )

# Step 2: FILTER out missing values and songs with zero popularity
# Songs with 0 popularity are usually unlisted or unreleased and would skew results
spotify_clean <- spotify_clean |>
  filter(
    !is.na(popularity),
    !is.na(loudness),
    !is.na(acousticness),
    !is.na(explicit),
    popularity > 0
  )

# Step 3: MUTATE to reformat the explicit variable into a clean labeled factor
# Changes True/False text into readable labels for the visualizations
spotify_clean <- spotify_clean |>
  mutate(
    explicit = case_when(
      explicit == "True"  ~ "Explicit",
      explicit == "False" ~ "Not Explicit"
    ),
    explicit = factor(explicit, levels = c("Not Explicit", "Explicit"))
  )

# Step 4: MUTATE to create acousticness groups
# Groups songs into Low, Medium, High so we can compare them easily
spotify_clean <- spotify_clean |>
  mutate(
    acousticness_level = case_when(
      acousticness < 0.33  ~ "Low Acousticness",
      acousticness < 0.66  ~ "Medium Acousticness",
      acousticness >= 0.66 ~ "High Acousticness"
    ),
    acousticness_level = factor(
      acousticness_level,
      levels = c("Low Acousticness", "Medium Acousticness", "High Acousticness")
    )
  )

# Step 5: DISTINCT to remove duplicate songs
# Keeps only one entry per unique song and artist combination
spotify_clean <- spotify_clean |>
  distinct(track_name, artists, .keep_all = TRUE)

# Step 6: SUMMARIZE average popularity by acousticness group
# Used for the bar chart in Visualization 3
acousticness_summary <- spotify_clean |>
  group_by(acousticness_level) |>
  summarize(
    avg_popularity = mean(popularity, na.rm = TRUE),
    count = n()
  ) |>
  arrange(desc(avg_popularity))

# Save the cleaned dataset
write_csv(spotify_clean, here("data", "spotify_clean.csv"))
```

To get the data ready for analysis, I went through several cleaning steps. First, I kept only the columns that were directly relevant to my three research questions and removed everything else like track IDs and album names. Then I filtered out songs with missing values or a popularity score of zero, since zero-popularity songs are usually not publicly available and would throw off the results. I reformatted the explicit column from plain text into a properly labeled category so it would display cleanly in visualizations. I also grouped the acousticness variable into three levels — Low, Medium, and High — since it is easier to compare groups than individual decimal numbers. Finally, I removed duplicate songs so each track only appears once, and calculated average popularity for each acousticness group to use in the bar chart.

---

## Question 1: Do Explicit Songs Tend to Be More Popular?

Some people assume that songs with explicit content attract more listeners because they tend to appear in hip-hop and pop — two of the most streamed genres on Spotify. This boxplot compares the full range of popularity scores between explicit and non-explicit songs, showing not just the averages but also how spread out the scores are for each group.

```{r}
#| label: fig-explicit-popularity
#| fig-cap: "Boxplot comparing popularity score distributions between explicit and non-explicit songs on Spotify."
#| fig-alt: "A side by side boxplot with two boxes. The left box represents non-explicit songs and the right box represents explicit songs. The x-axis shows the two groups and the y-axis shows popularity scores from 0 to 100. The median line inside each box shows the middle popularity score for each group. The explicit songs box has a slightly higher median than non-explicit songs. Both boxes have similar spread but explicit songs show a marginally higher overall popularity distribution. Outliers appear as small dots above and below each box."

spotify_clean |>
  ggplot(
    aes(
      x    = explicit,
      y    = popularity,
      fill = explicit
    )
  ) +
  geom_boxplot(
    outlier.shape = 21,
    outlier.size  = 1.2,
    outlier.alpha = 0.3,
    width         = 0.5,
    linewidth     = 0.4
  ) +
  scale_fill_manual(
    values = c(
      "Not Explicit" = "#5B8DB8",
      "Explicit"     = "#E07B54"
    )
  ) +
  labs(
    title    = "Are Explicit Songs More Popular on Spotify?",
    subtitle = "Comparing popularity score distributions for explicit vs non-explicit songs",
    x        = "Song Type",
    y        = "Popularity Score (0–100)",
    caption  = "Source: Spotify Tracks Dataset, Kaggle"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14),
    plot.subtitle      = element_text(color = "gray40", size = 11),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 12)
  )
```

Looking at the boxplot, explicit songs do tend to have a slightly higher median popularity score compared to non-explicit songs. However the difference is not dramatic — both groups have a wide spread of popularity scores, meaning there are both very popular and very unpopular songs in each category. This suggests that while explicit content may give a small boost to popularity, it is not the main driver of how well a song performs.

---

## Question 2: Do Louder Songs Tend to Be More Popular?

Loudness is one of the most direct audio features a producer controls. Songs that are louder tend to feel more energetic and punchy, which might make them more enjoyable to listeners. This scatterplot shows how each song's loudness level compares to its popularity score, with a trend line to show the overall direction of the relationship.

```{r}
#| label: fig-loudness-popularity
#| fig-cap: "Scatterplot showing the relationship between loudness and popularity for Spotify songs, with a trend line."
#| fig-alt: "A scatterplot where the x-axis shows loudness values ranging from about negative 50 decibels to 0 decibels, and the y-axis shows popularity scores from 0 to 100. Each point represents one song and is shown in a semi-transparent teal color to show where points overlap. A dark green trend line runs from the bottom left to the upper right of the plot, indicating that as loudness increases toward 0 decibels, popularity scores also tend to be higher. A shaded band around the trend line shows the uncertainty range. Most songs are clustered between negative 20 and 0 decibels."

spotify_clean |>
  ggplot(
    aes(
      x = loudness,
      y = popularity
    )
  ) +
  geom_point(
    alpha = 0.15,
    size  = 0.8,
    color = "#1D9E75"
  ) +
  geom_smooth(
    method    = "lm",
    se        = TRUE,
    color     = "#0F6E56",
    fill      = "#9FE1CB",
    linewidth = 1
  ) +
  labs(
    title    = "Do Louder Songs Get More Popular on Spotify?",
    subtitle = "Each point is one song — the line shows the overall trend",
    x        = "Loudness (decibels — closer to 0 means louder)",
    y        = "Popularity Score (0–100)",
    caption  = "Source: Spotify Tracks Dataset, Kaggle"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "gray40", size = 11),
    panel.grid.minor = element_blank()
  )
```

The scatterplot shows a clear positive trend — as loudness increases (gets closer to 0 decibels), popularity scores tend to go up as well. Most of the highly popular songs are clustered in the louder range between -10 and 0 decibels, while very quiet songs tend to have lower popularity scores. This makes sense because louder, more polished productions are more common in mainstream music, which naturally attracts more listeners on Spotify.

---

## Question 3: Are Acoustic Songs Less Popular Than Electronic Ones?

Acoustic songs tend to feel more raw and intimate, while heavily produced electronic songs are often designed for mass appeal. This bar chart compares the average popularity of songs grouped by how acoustic they sound — Low, Medium, or High acousticness — to see whether production style plays a role in how well a song performs on Spotify.

```{r}
#| label: fig-acousticness-popularity
#| fig-cap: "Bar chart comparing average Spotify popularity scores across Low, Medium, and High acousticness groups."
#| fig-alt: "A horizontal bar chart with three bars representing Low Acousticness, Medium Acousticness, and High Acousticness songs. The y-axis lists the three acousticness groups and the x-axis shows the average popularity score from 0 to 100. The bars are colored in shades of blue from light to dark. The Low Acousticness bar is the longest, indicating the highest average popularity, followed by Medium Acousticness, and then High Acousticness which has the shortest bar and lowest average popularity. Each bar has its exact average score labeled at the end."

acousticness_summary |>
  mutate(
    acousticness_level = fct_reorder(acousticness_level, avg_popularity)
  ) |>
  ggplot(
    aes(
      x    = avg_popularity,
      y    = acousticness_level,
      fill = avg_popularity
    )
  ) +
  geom_col(width = 0.55) +
  geom_text(
    aes(label = round(avg_popularity, 1)),
    hjust = -0.2,
    size  = 4,
    color = "gray20"
  ) +
  scale_fill_gradient(low = "#B5D4F4", high = "#0C447C") +
  scale_x_continuous(
    limits = c(0, 50),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title    = "Do Acoustic Songs Perform Worse on Spotify?",
    subtitle = "Average popularity score by acousticness level",
    x        = "Average Popularity Score (0–100)",
    y        = "Acousticness Level",
    caption  = "Source: Spotify Tracks Dataset, Kaggle"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14),
    plot.subtitle      = element_text(color = "gray40", size = 11),
    legend.position    = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 11)
  )
```

The bar chart clearly shows that songs with low acousticness — meaning they are more electronic and heavily produced — have the highest average popularity on Spotify. As acousticness increases, average popularity drops. This suggests that mainstream Spotify listeners tend to favor more produced, electronic-sounding music over raw acoustic recordings. This could also be because genres like pop, hip-hop, and electronic music — which dominate Spotify streams — tend to have very low acousticness scores.
