---
title: "Project: What Makes a Song Popular on Spotify?"
author: "Dhruv Salian"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Data Description

For this project, I used the Spotify Tracks Dataset from [Kaggle](https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset). This dataset contains information on over 114,000 songs available on Spotify, including audio characteristics and a popularity score assigned by the platform for each song.

Each row in the dataset represents one unique song. The key variables I used in this analysis are:

- **track_name**: The name of the song
- **artists**: Who performed the song
- **popularity**: A number from 0 to 100 showing how popular the song is on Spotify — higher means more popular
- **explicit**: Whether the song contains explicit content (True or False)
- **loudness**: How loud the song is, measured in decibels — most songs fall between -50 and 0
- **acousticness**: How acoustic the song sounds, from 0 (fully electronic) to 1 (fully acoustic)
- **track_genre**: The genre the song belongs to

The three questions I set out to answer were:

1. Do explicit songs tend to be more popular than non-explicit songs?
2. Do louder songs tend to be more popular?
3. Are acoustic songs less popular than electronic-sounding ones?

---

## Data Cleaning

```{r}
#| label: load-packages

library(tidyverse)
library(here)
```

```{r}
#| label: load-data

# Load the raw Spotify dataset using the here package
spotify_raw <- read_csv(here("data", "dataset.csv"))
```

```{r}
#| label: clean-data

# keeping only the columns i actually need
spotify_clean <- spotify_raw |>
  select(
    track_name,
    artists,
    track_genre,
    popularity,
    explicit,
    loudness,
    acousticness,
    danceability,
    energy,
    valence,
    tempo
  )

# dropping rows with missing values and removing zero popularity songs
# zero popularity means the song isnt really available to listeners
spotify_clean <- spotify_clean |>
  filter(
    !is.na(popularity),
    !is.na(loudness),
    !is.na(acousticness),
    !is.na(explicit),
    popularity > 0
  )

# converting explicit to readable labels
# handling both logical and text formats since the raw data can vary
spotify_clean <- spotify_clean |>
  mutate(
    explicit = case_when(
      explicit == TRUE  | explicit == "True"  ~ "Explicit",
      explicit == FALSE | explicit == "False" ~ "Not Explicit",
      TRUE ~ "Not Explicit"
    ),
    explicit = factor(explicit, levels = c("Not Explicit", "Explicit"))
  )

# bucketing acousticness into three groups so its easier to compare
spotify_clean <- spotify_clean |>
  mutate(
    acousticness_level = case_when(
      acousticness < 0.33  ~ "Low Acousticness",
      acousticness < 0.66  ~ "Medium Acousticness",
      acousticness >= 0.66 ~ "High Acousticness"
    ),
    acousticness_level = factor(
      acousticness_level,
      levels = c("Low Acousticness", "Medium Acousticness", "High Acousticness")
    )
  )

# removing duplicate tracks so each song only appears once
spotify_clean <- spotify_clean |>
  distinct(track_name, artists, .keep_all = TRUE)

# calculating average popularity per acousticness group for the bar chart
acousticness_summary <- spotify_clean |>
  group_by(acousticness_level) |>
  summarize(
    avg_popularity = mean(popularity, na.rm = TRUE),
    count = n()
  ) |>
  arrange(desc(avg_popularity))

# saving the cleaned data
write_csv(spotify_clean, here("data", "spotify_clean.csv"))
```

To get the data ready for analysis, I went through several cleaning steps. First, I kept only the columns that were directly relevant to my three research questions and removed everything else like track IDs and album names. Then I filtered out songs with missing values or a popularity score of zero since zero-popularity songs are usually not publicly available and would throw off the results. I reformatted the explicit column from plain text into a properly labeled category so it would display cleanly in visualizations. I also grouped the acousticness variable into three levels — Low, Medium, and High since it is easier to compare groups than individual decimal numbers. Finally, I removed duplicate songs so each track only appears once and calculated average popularity for each acousticness group to use in the bar chart.

---

## Question 1: Do Explicit Songs Tend to Be More Popular?

Some people assume that songs with explicit content attract more listeners because they tend to appear in hip-hop and pop which are two of the most streamed genres on Spotify. This violinplot compares the full range of popularity scores between explicit and non-explicit songs, showing not just the averages but also how spread out the scores are for each group.

```{r}
#| label: fig-explicit-popularity
#| fig-cap: "Violin plot comparing popularity score distributions between explicit and non-explicit songs on Spotify."
#| fig-alt: "A violin plot with two shapes side by side. The left violin represents non-explicit songs in blue and the right violin represents explicit songs in orange. The x-axis shows the two song types and the y-axis shows popularity scores from 0 to 100. The width of each violin at any point shows how many songs have that popularity score. The explicit songs violin is wider in the higher popularity range showing more songs with higher scores. A white dot inside each violin marks the median popularity score for that group."

spotify_clean |>
  ggplot(
    aes(
      x    = explicit,
      y    = popularity,
      fill = explicit
    )
  ) +
  geom_violin(
    trim        = TRUE,
    linewidth   = 0.4,
    alpha       = 0.85
  ) +
  geom_boxplot(
    width     = 0.08,
    fill      = "white",
    color     = "gray30",
    linewidth = 0.4,
    outlier.shape = NA
  ) +
  scale_fill_manual(
    values = c(
      "Not Explicit" = "#5B8DB8",
      "Explicit"     = "#E07B54"
    )
  ) +
  labs(
    title    = "Are Explicit Songs More Popular on Spotify?",
    subtitle = "Shape shows where most songs cluster — wider means more songs at that score",
    x        = "Song Type",
    y        = "Popularity Score (0–100)",
    caption  = "Source: Spotify Tracks Dataset, Kaggle"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14),
    plot.subtitle      = element_text(color = "gray40", size = 11),
    legend.position    = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(size = 12)
  )
```

Looking at the violin plot, explicit songs do tend to have a slightly higher median popularity score compared to non-explicit songs. The shape of each violin also shows that most songs in both groups cluster in the low to mid popularity range, with fewer songs reaching the very top. However the difference between the two groups is not dramatic, suggesting that while explicit content may give a small boost to popularity, it is not the main driver of how well a song performs.

---

## Question 2: Do Louder Songs Tend to Be More Popular?

Loudness is one of the most direct audio features a producer controls. Songs that are louder tend to feel more energetic and punchy, which might make them more enjoyable to listeners. To explore this I grouped all songs into four loudness ranges from very quiet all the way to very loud  and looked at how average popularity changes across those groups. The line chart makes it easy to see whether popularity steadily rises as songs get louder

```{r}
#| label: fig-loudness-popularity
#| fig-cap: "Line chart showing average Spotify popularity across four loudness groups, from very quiet to very loud."
#| fig-alt: "A line chart with four points connected by a line. The x-axis shows four loudness groups from left to right: -50 to -30 dB which is very quiet, -30 to -20 dB which is quiet, -20 to -10 dB which is loud, and -10 to 0 dB which is very loud. The y-axis shows average popularity score from 0 to 100. The line rises steadily from left to right showing that as songs get louder their average popularity increases. Each point on the line has the exact average score labeled above it. The points are colored in a gradient from light to dark green getting darker as loudness increases."

spotify_clean |>
  mutate(
    loudness_group = case_when(
      loudness < -30 ~ "-50 to -30 dB\n(Very Quiet)",
      loudness < -20 ~ "-30 to -20 dB\n(Quiet)",
      loudness < -10 ~ "-20 to -10 dB\n(Loud)",
      loudness >= -10 ~ "-10 to 0 dB\n(Very Loud)"
    ),
    loudness_group = factor(
      loudness_group,
      levels = c(
        "-50 to -30 dB\n(Very Quiet)",
        "-30 to -20 dB\n(Quiet)",
        "-20 to -10 dB\n(Loud)",
        "-10 to 0 dB\n(Very Loud)"
      )
    )
  ) |>
  group_by(loudness_group) |>
  summarize(
    avg_popularity = mean(popularity, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(
    aes(
      x = loudness_group,
      y = avg_popularity,
      group = 1
    )
  ) +
  geom_line(
    color     = "#1D9E75",
    linewidth = 1.2
  ) +
  geom_point(
    aes(color = avg_popularity),
    size = 5
  ) +
  geom_text(
    aes(label = round(avg_popularity, 1)),
    vjust    = -1.2,
    size     = 4,
    fontface = "bold",
    color    = "gray20"
  ) +
  scale_color_gradient(low = "#9FE1CB", high = "#085041") +
  scale_y_continuous(
    limits = c(20, 55),
    expand = expansion(mult = c(0.05, 0.1))
  ) +
  labs(
    title    = "Do Louder Songs Get More Popular on Spotify?",
    subtitle = "Average popularity score rises as songs get louder",
    x        = "Loudness Group",
    y        = "Average Popularity Score (0–100)",
    caption  = "Source: Spotify Tracks Dataset, Kaggle"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "gray40", size = 11),
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 10, lineheight = 1.3)
  )
```

The line chart shows a clear upward trend as songs get louder the average popularity score rises steadily across all four groups. The jump is most noticeable between the quieter groups and the louder ones with very loud songs scoring noticeably higher on average. This makes sense because louder more polished productions are more common in mainstream music which naturally attracts more listeners on Spotify.

---

## Question 3: Are Acoustic Songs Less Popular Than Electronic Ones?

Acoustic songs tend to feel more raw and intimate while heavily produced electronic songs are often designed for mass appeal. This bar chart compares the average popularity of songs grouped by how acoustic they sound — Low, Medium, or High acousticness  to see whether production style plays a role in how well a song performs on Spotify.

```{r}
#| label: fig-acousticness-popularity
#| fig-cap: "Bar chart comparing average Spotify popularity scores across Low, Medium, and High acousticness groups."
#| fig-alt: "A horizontal bar chart with three bars representing Low Acousticness, Medium Acousticness, and High Acousticness songs. The y-axis lists the three acousticness groups and the x-axis shows the average popularity score from 0 to 100. The bars are colored in shades of blue from light to dark. The Low Acousticness bar is the longest, indicating the highest average popularity, followed by Medium Acousticness, and then High Acousticness which has the shortest bar and lowest average popularity. Each bar has its exact average score labeled at the end."

acousticness_summary |>
  mutate(
    acousticness_level = fct_reorder(acousticness_level, avg_popularity)
  ) |>
  ggplot(
    aes(
      x    = avg_popularity,
      y    = acousticness_level,
      fill = avg_popularity
    )
  ) +
  geom_col(width = 0.55) +
  geom_text(
    aes(label = round(avg_popularity, 1)),
    hjust = -0.2,
    size  = 4,
    color = "gray20"
  ) +
  scale_fill_gradient(low = "#B5D4F4", high = "#0C447C") +
  scale_x_continuous(
    limits = c(0, 50),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title    = "Do Acoustic Songs Perform Worse on Spotify?",
    subtitle = "Average popularity score by acousticness level",
    x        = "Average Popularity Score (0–100)",
    y        = "Acousticness Level",
    caption  = "Source: Spotify Tracks Dataset, Kaggle"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14),
    plot.subtitle      = element_text(color = "gray40", size = 11),
    legend.position    = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 11)
  )
```

The bar chart clearly shows that songs with low acousticness  meaning they are more electronic and heavily produced  have the highest average popularity on Spotify. As acousticness increases average popularity drops. This suggests that mainstream Spotify listeners tend to favor more produced electronic sounding music over raw acoustic recordings. This could also be because genres like pop, hip-hop, and electronic music  which dominate Spotify streams.